---
title: "ReproducibleResearch_PeerReview01"
author: "Craig Rowley"
date: "December 6, 2014"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

I have something I want to show you today that I think you'll enjoy.  And it will benefit your understanding of quantified health.

Let's start by loading our activity dataset (you'll need to set your working directory appropriately):

```{r}
activity <- read.csv("activity.csv", header=TRUE ,sep=",")
rowCount <- nrow(activity)

```

`r rowCount` rows of beautiful data!

Let's see what we're dealing with:

```{r}
summary(activity)

```


```{r echo=FALSE,}
naStepsCount <- nrow( activity[is.na(activity$steps),] )
pctNaStepsCount <- round( 100*naStepsCount/rowCount, 2 )

zeroStepsCount <- nrow( activity[activity$steps == 0 & !is.na(activity$steps),] )
pctZeroStepsCount <- round( 100*zeroStepsCount/rowCount, 2 )

```

So:  

- steps is NA `r pctNaStepsCount`% of the time, and `r pctZeroStepsCount`% of the populated values are 0  
- date summary just gives us a view of the data... it is an ISO formatted Date
- interval summary is also odd looking in the summary

---

### Steps

TBD :: We need to set an approach to impute steps... set to zero?



---

### Date

Let's factorize date and look at that again:

```{r}
activity$date <- factor(activity$date)
summary(activity$date)
```


There we go!  Now we see the Date range from Oct 1st through Nov 30th.

---

### Interval

Let's look at interval now:

```{r}
head(activity$interval)
```
```{r}
tail(activity$interval)
```

Odd, so we know it is an identifier, so it can be factored and we can look at unique values:

```{r}
activity$interval <- factor(activity$interval)
levels(activity$interval)
```

```{r echo=FALSE}
lenIntervalLevels <- length( levels(activity$interval))
```

There are `r lenIntervalLevels` different values, which corresponds exactly to the number of 5-minute intervals in a day. Examining the levels closer, it looks like the last two digits range from zero to 55, and the first digits range from non-existant to 23.  24-hour clock has a similar range, so let's assume these are clock times:  00:00 through 23:55.









